Executive Constraints and Civil Conflict Onset

### 2020-08-30 paper.Rmd
Finish alternative specifications & update R dependencies

---
title: Executive Constraints and Civil Conflict Onset
author:
    name: J.L.A. Krusell
    affiliation: Department of Peace and Conflict Research
    supervisor: Kristine Eck
    degree: "Master's Thesis"
date: "Spring 2020"
titlelogo: |
    ```{=latex}
    assets/UU_logo_CMYK-eps-converted-to.pdf
    ```
wordcount: 14000
suppress-bibliography: false
bibliography: library.bib
csl: assets/chicago-author-date.csl
link-citations: true
documentclass: scrreprt
classoption: paper=letter
geometry:
    - top=30mm
    - heightrounded
fontsize: 12pt
linestretch: 1.5
header-includes:
    - \usepackage{amsmath}
    - \usepackage{etoolbox}
    - \usepackage{lualatex-math}
    - \usepackage{unicode-math}
    - \appto\Shaded{\singlespacing}
    - \BeforeBeginEnvironment{align*}{\begin{singlespace}\vspace*{-\baselineskip}}
    - \AfterEndEnvironment{align*}{\end{singlespace}\noindent\ignorespaces}
    - \counterwithout{table}{chapter}
    - \counterwithout{figure}{chapter}
    - \usepackage[font={small}]{caption}
output:
    pdf_document:
        fig_caption: true
        keep_tex: true
        latex_engine: lualatex
        toc: true
        toc_depth: 2
        pandoc_args: ["--template", "assets/thesis_template.latex",
                      "--syntax-definition", "assets/stan.xml",
                      "--highlight-style", "kate",
                      "--lua-filter", "utils/text_refs/text_refs.lua"]
    html_document:
        smart: true
        toc: true
abstract: Lorem ipsum
---

@@

options(knitr.kable.NA = ".")
opts_chunk$set(echo = F, message = F, warning = F,
               fig.width = 5, fig.height = 3, cache = T)

# TODO: this has been added to package
fmt <- function(x, digits = 2) {
    prettyNum(round(x, digits), big.mark = ",")
}

# Create initial descriptive stats for introduction
ucdp <- readRDS("./data/raw/UcdpPrioConflict_v19_1.rds")

@@

In the UCDP/PRIO armed conflict dataset, high intensity is defined at
the threshold of 1,000 battle deaths. High intensity conflicts, or
civil wars, can be coded according to when the threshold for high
intensity is crosses either by a single year or by the cumulative
battle deaths.[^battle_deaths] I therefore re-run the same model,
including the same priors and runtime parameters, twice for the two
different operational definitions. In both cases, I retain an episodic
delineation between separate conflicts --- in other words, conflicts
over the same incompatibility can have multiple onsets when there are
lulls in fighting.[^episode_gaps] For the former, the onset year is
set as the first year which crosses the high intensity threshold for
each period of sustained fighting of at least 25 battle deaths. For
the latter, onset is the coded conflict year for the cumulative period
passing the high intensity threshold.

@@

args <- list(pars = "beta",
             legend_labels = c("Low & High Intensity",
                               "High Intensity",
                               "Cumulatively High Intensity"))
do.call(overlay_pars, c(ll, args)) +
    scale_x_discrete(labels = colnames(beta))
```

@@

    # Really could just grab this from the existing `beta` variable,
    # but just to be safe...
    load(file.path("posteriors", models[i], "input.RData"),
         envir = e <- new.env())
    colnames(beta) <- c("exec_constraints", colnames(e$X),
                        sprintf("exec * %s",
                                colnames(e$X)[e$interaction_idx]))

@@

Figure \ref{fig:overlay} plots the original results compared to the
results from running the model with the two different operational
definitions of high intensity conflict.[^theta] The majority of
depicted parameter estimates are fairly stable across runs. Notably, the relationship between executive constraints and conflict onsets remains negative with a high posterior probability of being less than zero. The estimates for the interaction effect between executive constraints and GDP per capita, however, become far to uncertain to be able to conclusively assert an association. The probability that the interaction term is less than zero for the high intensity model run, *i.e.* where at least a single year crosses the high intensity threshold, is only `r fmt(ix_probs["major_only"])` while the probability for the cumulative intensity model is only
`r fmt(ix_probs["cumulative_only"])`.

@@

(ref:confusion_major) Confusion matrix for conflicts with at least a
    single year passing the high intensity threshold with a 0.168
    cutoff for the median posterior predicted probabilities.

```{r confusion_major}
extra <- read_post("./posteriors/major_only/extra_posteriors.csv")

@@

load("./posteriors/major_only/input.RData")

m <- table(p_median > .168, final.df$lmajor_onset)
kable(m, "latex", align = "c", booktabs = T,
      linesep = "", escape = F, caption = "hello")
```

(ref:confusion_cumulative) Confusion matrix for conflicts with a
    cumulative high intensity with a 0.168 cutoff for the median
    posterior predicted probabilities.

```{r confusion_cumulative}
extra <- read_post("./posteriors/cumulative_only/extra_posteriors.csv")

@@

load("./posteriors/cumulative_only/input.RData")

m <- table(p_median > .168, final.df$lcum_onset)
kable(m, "latex", align = "c", booktabs = T,
      linesep = "", escape = F,
      caption = "world")
```

How well does the model fit the data when only examining high intensity conflicts?

>>>>

---
title: Executive Constraints and Civil Conflict Onset
author:
    name: J.L.A. Krusell
    affiliation: Department of Peace and Conflict Research
    supervisor: Kristine Eck
    degree: "Master's Thesis"
date: "Spring 2020"
titlelogo: |
    ```{=latex}
    assets/UU_logo_CMYK-eps-converted-to.pdf
    ```
wordcount: 14300
suppress-bibliography: false
bibliography: library.bib
csl: assets/chicago-author-date.csl
link-citations: true
documentclass: scrreprt
classoption: paper=letter
geometry:
    - top=30mm
    - heightrounded
fontsize: 12pt
linestretch: 1.5
header-includes:
    - \usepackage{amsmath}
    - \usepackage{etoolbox}
    - \usepackage{lualatex-math}
    - \usepackage{unicode-math}
    - \appto\Shaded{\singlespacing}
    - \BeforeBeginEnvironment{align*}{\begin{singlespace}\vspace*{-\baselineskip}}
    - \AfterEndEnvironment{align*}{\end{singlespace}\noindent\ignorespaces}
    - \counterwithout{table}{chapter}
    - \counterwithout{figure}{chapter}
    - \usepackage[font={small}]{caption}
output:
    pdf_document:
        fig_caption: true
        keep_tex: true
        latex_engine: lualatex
        toc: true
        toc_depth: 2
        pandoc_args: ["--template", "assets/thesis_template.latex",
                      "--syntax-definition", "assets/stan.xml",
                      "--highlight-style", "kate",
                      "--lua-filter", "utils/text_refs/text_refs.lua"]
    html_document:
        smart: true
        toc: true
abstract: Lorem ipsum
---

@@

options(knitr.kable.NA = ".")
opts_chunk$set(echo = F, message = F, warning = F,
               fig.width = 5, fig.height = 3, cache = T)
```

```{r intro-descriptive-stats}
# Create initial descriptive stats for introduction
ucdp <- readRDS("./data/raw/UcdpPrioConflict_v19_1.rds")

@@

```{r}
load("./posteriors/cumulative_only/input.RData",
     envir = cumulative <- new.env())

load("./posteriors/major_only/input.RData",
     envir = major <- new.env())
```

In the UCDP/PRIO armed conflict dataset, high intensity is defined at
the threshold of 1,000 battle deaths. High intensity conflicts, or
civil wars, can be coded according to when the threshold for high
intensity is crossed either by a single year or by the cumulative
battle death count.[^battle_deaths] I therefore re-run the same model,
including the same priors and runtime parameters, twice for the two
different operational definitions. In both cases, I retain an episodic
delineation between separate conflicts --- in other words, conflicts
over the same incompatibility can have multiple onsets when there are
lulls in fighting.[^episode_gaps] For the former, the onset year is
set as the first year which crosses the high intensity threshold for
each period of sustained fighting of at least 25 battle deaths. For
the latter, onset is the first coded conflict year for the cumulative
period passing the high intensity threshold. In total there
`r sum(major$final.df$lmajor_onset)` onsets for the single year 1,000
battle deaths definition, and `r sum(cumulative$final.df$lcum_onset)`
onsets for the cumulative definition.

@@

args <- list(pars = "beta",
             legend_labels = c("Low & High Intensity",
                               "High Intensity",
                               "Cumulative High Intensity"))
do.call(overlay_pars, c(ll, args)) +
    scale_x_discrete(labels = colnames(beta))
```

@@

    if (models[i] == "major_only") {
        colnames(beta) <- c("exec_constraints", colnames(major$X),
                            sprintf("exec * %s",
                                    colnames(major$X)[major$interaction_idx]))
    } else {
        colnames(beta) <- c("exec_constraints", colnames(cumulative$X),
                            sprintf("exec * %s",
                                    colnames(cumulative$X)[cumulative$interaction_idx]))
    }

@@

Figure \ref{fig:overlay} plots the original regression slopes compared
to the results from running the model with the two different
operational definitions of high intensity conflict.[^theta] The
majority of depicted parameter estimates are fairly stable across
runs. Notably, the relationship between executive constraints and
conflict onsets remains negative with a high posterior probability of
being less than zero. The estimates for the interaction effect between
executive constraints and GDP per capita, however, become far too
uncertain to be able to conclusively assert an association. The
probability that the interaction term is less than zero for the high
intensity model run, *i.e.* where at least a single year crosses the
high intensity threshold, is only `r fmt(ix_probs["major_only"])`
while the probability for the cumulative intensity model run is only
`r fmt(ix_probs["cumulative_only"])`.

@@

(ref:confusion-major) Confusion matrix with a 0.168 cutoff for the
    median posterior predicted probabilities from the single year high
    intensity model run.

```{r confusion-major}
extra <- read_post("./posteriors/major_only/extra_posteriors.csv")

@@

m <- table(p_median > .168, major$final.df$lmajor_onset)
rownames(m) <- c("Predicted: No Onset", "Predicted: Onset")
colnames(m) <- c("Actual: No Onset", "Actual: Onset")

# Calculate the AUCs as well
x <- evalmod(scores = p_median, labels = major$final.df$lmajor_onset)
major_auc <- attr(x$roc[[1]], "auc") %>% fmt
major_auc_pr <- attr(x$prcs[[1]], "auc") %>% fmt

kable(m, "latex", align = "c", booktabs = T,
      linesep = "", escape = F, caption = "(ref:confusion-major)")
```

Table \ref{tab:confusion-major} presents the confusion matrix for the
single year high intensity model run using the median posterior
predicted probabilities. A cutoff of 0.168 was reused for comparative
purposes to the original results. In this case, the model is unable to
predict even a single major conflict onset with the highest median
posterior predicted probability for an actual onset being only
`r fmt(max(p_median[major$final.df$lmajor_onset == 1]))`. Furthermore,
although the ROC AUC is `r major_auc`, this is again due to the major
class imbalance between positive versus negative outcomes in the
dataset. A precision-recall AUC of `r major_auc_pr` demonstrates the
poor performance of the model.

(ref:confusion-cumulative) Confusion matrix with a 0.168 cutoff for
    the median posterior predicted probabilities from the cumulative
    high intensity model run.

```{r confusion-cumulative}
extra <- read_post("./posteriors/cumulative_only/extra_posteriors.csv")

@@

m <- table(p_median > .168, cumulative$final.df$lcum_onset)
rownames(m) <- c("Predicted: No Onset", "Predicted: Onset")
colnames(m) <- c("Actual: No Onset", "Actual: Onset")

x <- evalmod(scores = p_median, labels = cumulative$final.df$lcum_onset)
cum_auc <- attr(x$roc[[1]], "auc") %>% fmt
cum_auc_pr <- attr(x$prcs[[1]], "auc") %>% fmt

kable(m, "latex", align = "c", booktabs = T,
      linesep = "", escape = F,
      caption = "(ref:confusion-cumulative)")
```

Similarly, table \ref{tab:confusion-cumulative} shows the confusion
matrix for the cumulative high intensity model run with the same
median posterior predicted probability cutoff. Although the model in
this case is able to correctly classify 17 percent of onsets with an
ROC AUC of `r cum_auc`, the overall performance is fairly limited with
a precision-recall AUC of only `r cum_auc_pr`.

Taken together the model runs examining high intensity conflicts
suggest that the negative relationship between executive constraints
and civil conflict onset is fairly robust. However, the same issues of
model fit continue to plague
